### 2013-08-01 Doc/MMVR2014/method.tex
ADD: new section about the methods showing the combination of matrices
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@607 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

% \subsection{Biomechanical Models and Coupling}
It is known that the parenchyma exhibits non-linear viscoelastic behavior \cite{Marchesseau2010}.
However, as we are mainly interested in the static equilibrium, we do not model the time-dependent
phenomena related to viscosity. In the simulation we employ corotational finite elements~\cite{Felippa2005}.
Although it relies on linear stress-strain relationship, large displacements including rotations are modeled correctly.
% Briefly, the motion of each element $e$ is decomposed into rigid rotation $\Mat{R}^e$ and local deformation in each step of the simulation. 
% The rotations are then used to update each local element stiffness matrix as $\Mat{R}^e\Mat{K}_0{\Mat{R}^e}^{\top}$
% whereas the deformations are used to compute the linear strain in the local corotational frame.
% There are several ways of computing the corotational frame for elements; we rely on
% the geometrical method proposed in \cite{Nesme2005}.

@@

% \subsection{Glisson's Capsule} %{{{
% \label{ss:capsuleModel}
The thickness of the Glisson's capsule is relatively small: the values in range of 10-20
$\mu$m have been reported in~\cite{Umale2011}.
It is not possible to model such thin structure with classical tetrahedral
elements:
%if the real-time aspect of the simulation is to be preserved.
%Furthermore, modeling both the tissue and the capsule 
the model would require an extremely 
dense mesh to avoid numerical instabilities and thus would significantly
violate the speed requirements imposed for medical simulators.
Instead, modeling the capsule with two-dimensional elements that abstract from the
thickness in the third dimension seems
to be a natural choice. In the elasticity theory, this functionality is usually provided by membrane and shell elements.
Based on the observation of its behaviour, we also
assume negligible bending forces and propose a model based on membrane
elements. 
To maintain simplicity of the composite model we choose simple triangular
elements with constant strain (CST).

The computation of elastic stiffness matrix follows the common derivation
%
\begin{eqnarray}
  \Mat{K}^m & = & \int_V \Mat{B}^T \Mat{E} \Mat{B} dV     \label{mem1} \\
            & = & h \int_A \Mat{B}^T \Mat{E} \Mat{B} dA   \label{mem2} \\
            & = & h A \Mat{B}^T \Mat{E} \Mat{B}           \label{mem3}
\end{eqnarray}
%
where $\Mat{B}$ is the strain-displacement matrix, $\Mat{E}$ the material
matrix, $h$ is the thickness and $A$ area of the element. In the previous
\eqref{mem2} follows from the fact that we assume constant thickness of the
element and \eqref{mem3} follows from the fact that the strain-displacement
matrix is constant in our case. The strain-displacement matrix for the CST
element can be expressed as:
%
\begin{equation}
  \Mat{B} = \frac{1}{2A} \begin{bmatrix}
    y_{23} & 0      & y_{31} & 0      & y_{12} & 0 \\
         0 & x_{32} & 0      & x_{13} & 0      & x_{21} \\
    x_{32} & y_{23} & x_{13} & y_{31} & x_{21} & y_{12}
  \end{bmatrix}
\end{equation}
%
The values $x_{ij} = x_i - x_j$ and $y_{ij} = y_i - y_j$ are computed from
the $x$ or $y$ coordinates of the nodes $i,j$ of the triangular element.
The reader can refer to the respective literature~\cite{Felippa2003} for more thorough
description.

Similarly as with model of parenchyma we use linear elastic material and employ
the corotational formulation for the CST elements.

@@

%\subsection{Mechanical Coupling between Parenchyma and Capsule} %{{{
The literature reports high cohesion between capsule and parenchyma.
Based on this property we assume there is no relative motion of the capsule \wrt\ the parenchyma.
Although an arbitrary surface mesh could be used to model the capsule, we exploit 
the fact that the parenchyma is modeled by tetrahedral elements having
triangular faces. Thus, as the boundary of the volumetric mesh is already
triangulated, we simply employ the triangles on the mesh surface to model the capsule.

Using directly the boundary of the tetrahedral mesh does not only solve the
problem of building the surface mesh, but has one more advantage: the nodes
of the triangular mesh coincide with the nodes of the tetrahedral mesh, so no projection of one mesh onto the other is needed.
and the stiffness matrices for capsule and parenchyma are then easily assembled together by adding the mechanical contribution 
of a triangle to the corresponding tetrahedron.
% and solved as one system.
%
%\CD{The following may appear as "trivial"... maybe we can remove this part at the end if we need space} 
Without the loss of generality we can assume the tetrahedron consists of
nodes $p_1, p_2, p_3$ and $p_4$ and the boundary triangle has nodes $p_1, p_2$
and $p_3$. We can reorder the degrees of freedom so that the stiffness
matrix $\Mat{K}^t$ for the tetrahedron can be written as:
%
\begin{equation}
  \Mat{K}^t = \left[\begin{array}{c|c}
      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
      \hline
      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\.
  \end{array}\right]
\end{equation}
%
Hence, the assembled stiffness matrix for the element is
%
\begin{equation}
  \Mat{K} = \left[\begin{array}{c|c}
      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
      \hline
      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
  \end{array}\right]
  +
  \left[\begin{array}{c|c}
      \Mat{K}^m & 0 \\
      \hline
      0 & 0 \\
  \end{array}\right]
\end{equation}
%
where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.

The resulting system of linear equations is solved by direct solver based on Cholesky decomposition.

>>>>

% \subsection{Biomechanical Models and Coupling}
%It is known that the parenchyma exhibits non-linear viscoelastic behavior \cite{Marchesseau2010}.
Since in this paper, we are mainly interested in the static equilibrium, we do not model the time-dependent
phenomena related to viscosity. In the simulation we employ corotational finite elements;
There are several ways of computing the corotational frame for elements; we rely on the geometrical method proposed in~\cite{Nesme2005}.
%although it relies on linear stress-strain relationship, large displacements including rotations are modeled correctly.
Denoting $p$ a generic tetrahedral element of the parenchyma, the 12$\times$12 element stiffness matrix is then computed as
\begin{equation}
\Mat{K}_p = \int_{V_p} \Mat{B}_p^\top \Mat{D}_p \Mat{B}_p dV
\end{equation}
where $\Mat{B}_p$ is the strain-displacement matrix and $\Mat{D}_p$ is the stress-strain matrix.

While the thickness of a vessel wall does not exceed 250$\mu$m~\cite{Forauer2003}, stiffness of 0.62$\pm$0.41\,MPa has been reported 
in~\cite{Umale2011}, being quite important when compared to the parenchyma. Due to this property, vessels influence 
the mechanical response of the tissue as illustrated in~\cite{Peterlik2012}. Since the vessel wall cannot be modeled 
with tetrahedral elements if the real-time aspect of the simulation is to be preserved, 
serially-linked beam elements based on Tymoshenko formulation are employed, taking into account the hollow 
structure of the vessels via moments of inertia. 
%to model the vessels which allows for correct simulation of large deformations as it is based on the corotational framework 
%similar to the one used for the parenchyma. 
For an arbitrary beam element representing a segment of a vessel, the corresponding local stiffness matrix $\Mat{K}_v$ is computed as shown in~\cite{Duriez2006}.

% Briefly, the motion of each element $e$ is decomposed into rigid rotation $\Mat{R}^e$ and local deformation in each step of the simulation. 
% The rotations are then used to update each local element stiffness matrix as $\Mat{R}^e\Mat{K}_0{\Mat{R}^e}^{\top}$
% whereas the deformations are used to compute the linear strain in the local corotational frame.
% There are several ways of computing the corotational frame for elements; we rely on
% the geometrical method proposed in \cite{Nesme2005}.

@@

% \subsection{Glisson's Capsule} %{{{
% \label{ss:capsuleModel}
The thickness of the Glisson's capsule is also very small: the values in range of 10-20 $\mu$m have been reported in~\cite{Umale2011}.
It is not possible to model such thin structure with classical tetrahedral
elements:
%if the real-time aspect of the simulation is to be preserved.
%Furthermore, modeling both the tissue and the capsule 
the model would require an extremely 
dense mesh to avoid numerical instabilities and thus would significantly
violate the speed requirements imposed for medical simulators.
Instead, modeling the capsule with two-dimensional elements that abstract from the
thickness in the third dimension seems
to be a natural choice. In the elasticity theory, this functionality is usually provided by membrane elements.
%Based on the observation of its behaviour, we also
%assume negligible bending forces and propose a model based on membrane
%elements. 
To maintain simplicity of the composite model we choose simple triangular
elements with constant strain (CST).

The 9$\times$9 stiffness matrix $\Mat{K}_m$ is given as 
%
\begin{equation}
 \Mat{K}_c = \int_{S_c} \Mat{B}_c^\top \Mat{D}_c \Mat{B}_c dS 
\end{equation}
where $\Mat{B}_c$ and $\Mat{D}_c$ are strain-displacement and strain-stress matrices, respectively.

The complete model of the liver takes into account all three components described above by combining the contribution 
of parenchyma, vessels and capsule. The coupling between the vessel elements (beams) and parenchyma elements (tetrahedra)
is described in detail in~\cite{Peterlik2012}. Briefly, each vessel node (having 6 degrees of freedom to account for torques in vessels) 
is coupled with a tetrahedra in which it is located via barycentric coordinates. This coupling remains constant during 
the simulation and can be described via matrix $\Mat{J}_{v\rightarrow p}$ which is a 12$\times$6 Jacobian matrix of the coupling.

The coupling between the parenchyma and capsule is straightforward, since the triangles used as the 
domain for the CST formulations are the surface faces of the volume mesh. Therefore, for given triangle 
with vertices $v_1$, $v_2$ and $v_3$, the corresponding tetrahedron (sharing the same vertices is found) and 
a 9$\times$12 permutation matrix $\Mat{P}_{c\rightarrow p}$ mapping the triangle vertices to tetrahedra vertices is computed.

Without loss of generality, let us suppose a tetrahedral element $e$ 
which receives a mechanical contribution from both the capsule and vessels. The composite element stiffness matrix $\Mat{K}_e$
is then computed as:
\begin{equation}
\Mat{K}_e = \Mat{K}_p + \Mat{J}_{v\rightarrow p} \Mat{K}_v \Mat{J}^\top_{v\rightarrow p} + \Mat{P}_{c\rightarrow p} \Mat{K}_c \Mat{P}_{c\rightarrow p}^\top.
\end{equation}

% 
% \begin{eqnarray}
%   \Mat{K}^m & = & \int_V \Mat{B}^T \Mat{E} \Mat{B} dV     \label{mem1} \\
%             & = & h \int_A \Mat{B}^T \Mat{E} \Mat{B} dA   \label{mem2} \\
%             & = & h A \Mat{B}^T \Mat{E} \Mat{B}           \label{mem3}
% \end{eqnarray}
% %
% where $\Mat{B}$ is the strain-displacement matrix, $\Mat{E}$ the material
% matrix, $h$ is the thickness and $A$ area of the element. In the previous
% \eqref{mem2} follows from the fact that we assume constant thickness of the
% element and \eqref{mem3} follows from the fact that the strain-displacement
% matrix is constant in our case. The strain-displacement matrix for the CST
% element can be expressed as:
% %
% \begin{equation}
%   \Mat{B} = \frac{1}{2A} \begin{bmatrix}
%     y_{23} & 0      & y_{31} & 0      & y_{12} & 0 \\
%          0 & x_{32} & 0      & x_{13} & 0      & x_{21} \\
%     x_{32} & y_{23} & x_{13} & y_{31} & x_{21} & y_{12}
%   \end{bmatrix}
% \end{equation}
% %
% The values $x_{ij} = x_i - x_j$ and $y_{ij} = y_i - y_j$ are computed from
% the $x$ or $y$ coordinates of the nodes $i,j$ of the triangular element.
% The reader can refer to the respective literature~\cite{Felippa2003} for more thorough
% description.
% 
% Similarly as with model of parenchyma we use linear elastic material and employ
% the corotational formulation for the CST elements.

@@

%\subsection{Mechanical Coupling between Parenchyma and Capsule} %{{{
% The literature reports high cohesion between capsule and parenchyma.
% Based on this property we assume there is no relative motion of the capsule \wrt\ the parenchyma.
% Although an arbitrary surface mesh could be used to model the capsule, we exploit 
% the fact that the parenchyma is modeled by tetrahedral elements having
% triangular faces. Thus, as the boundary of the volumetric mesh is already
% triangulated, we simply employ the triangles on the mesh surface to model the capsule.
% 
% Using directly the boundary of the tetrahedral mesh does not only solve the
% problem of building the surface mesh, but has one more advantage: the nodes
% of the triangular mesh coincide with the nodes of the tetrahedral mesh, so no projection of one mesh onto the other is needed.
% and the stiffness matrices for capsule and parenchyma are then easily assembled together by adding the mechanical contribution 
% of a triangle to the corresponding tetrahedron.
% % and solved as one system.
% %
% %\CD{The following may appear as "trivial"... maybe we can remove this part at the end if we need space} 
% Without the loss of generality we can assume the tetrahedron consists of
% nodes $p_1, p_2, p_3$ and $p_4$ and the boundary triangle has nodes $p_1, p_2$
% and $p_3$. We can reorder the degrees of freedom so that the stiffness
% matrix $\Mat{K}^t$ for the tetrahedron can be written as:
% %
% \begin{equation}
%   \Mat{K}^t = \left[\begin{array}{c|c}
%       \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
%       \hline
%       \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\.
%   \end{array}\right]
% \end{equation}
% %
% Hence, the assembled stiffness matrix for the element is
% %
% \begin{equation}
%   \Mat{K} = \left[\begin{array}{c|c}
%       \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
%       \hline
%       \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
%   \end{array}\right]
%   +
%   \left[\begin{array}{c|c}
%       \Mat{K}^m & 0 \\
%       \hline
%       0 & 0 \\
%   \end{array}\right]
% \end{equation}
% %
% where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.
% 
% The resulting system of linear equations is solved by direct solver based on Cholesky decomposition.
